The abundance of Bullet-Groups in $\Lambda$CDM

### 2014-03-11 paper/ms.tex
conclusions

As a zero-th order approximation, in this paper we work with three
quantities that are available from observations of Bullet-like
systems. The projected distance between two dominant DM clumps, the
projected distance between the DM-baryonic clumps and the ratio of the galaxies
luminosities associated to the DM clumps.

From the simulation point of view, the first quantity can be
translated into the 2D projected values of $||\vec{r}||$ and its value
relative to the virial radius $D_{\rm   off}= ||\vec{r}||_{\rm
  2D}/R_{\rm vir}$. The last quantity can be approximated by ratio of
the circular velocities of the two clumps $V_{\rm c,sub}/V_{\rm
  c,host}$ which under conditions of constant mass-to-light ratios
should be equal to the square root of the luminosity ratio. The second
quantity is not directly available from a DM-only simulation but, as we
show in the Results section, can be  inferred from the available
information.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.9\textwidth]{figure_3.eps}
\end{center}
\caption{Integrated probability distribution for the relative velocity of the
sub-halo with respect to its host. The left panel shows the results
in physical units while the right panel show the same values
normalized by the circular velocity of the host halo. The line coding
follows the same structure as Figure \ref{fig:displacement}}
\label{fig:velocities}
\end{figure*}

In Figure \ref{fig:velocities} we present the integrated probability
of the relative peculiar velocities of the sub-halos with respect to
the host halo. The left panel presents this velocity in physical units
while the left panel presents them as a fraction of the circular
velocity of the host halo.

The panel with the normalized velocities shows that the distribution
of sub-halo velocities is close to universal. Regardless of the mass
of the host halo and the redshift the integrated distributions lie
very close to each other.

@@

Figure \ref{fig:geometry} presents the geometry of the bullet groups using the 
variables $\mu$ and $D_{\rm off}$. The first evident feature is that
most of the configurations have $|mu|>0.9$ ($\theta\leq 30^{\circ}$),
meaning that most of the collisions can be described as a head-on while
only a minority with $|mu|<0.9$ have more tangential trajectories. For
the pairs on radial trajectories there are three regions of interest
in this plane that describe different stages in the collision,
assuming that the sub-halo merges or falls below the detection
threshold after the first pass through the center of the host halo.

The first region has $\mu\approx-1$ and $D_{\rm off}>0.6$, which
locates the systems where a head-on collision has just started. The
sub-halo is close to the boundary of the host halo and is
infalling. The second region has $\mu\approx 1$ and $D_{\rm off}<0.6$;
at this stage the collision continues after the first crossing of the
host's center, the low number of halos with radial infalling velocities
and displacements $D_{\rm off}>0.6$ suggest that this the maximum
range of radii for the apogee.  The third region corresponds to
$\mu\approx-1$ and $D_{\rm off}<0.6$ which corresponds to the
secondary infall after apogee.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{figure_5.eps}
\end{center}
\caption{Integrated probability distribution for the estimated
  baryonic displacements in the group and cluster samples. The
  continuous line marked as $P_{2D}$ shows the statistics reported by
  \citet{ForeroRomero2010} for a cosmological simulation including DM
  and baryons. The vertical lines correspond to the mean value and
  uncertainty of the displacement measured for \bullg.}
\label{fig:baryonic_displacements}
\end{figure}

These different collision geometries will produce different results in
terms of the displacement between dark matter and baryons. Strictly
speaking, the results we have derived so far, apply to multi-modal groups and their
expected separation between the two dominant dark matter clumps but it
should not be interpreted that all have a corresponding DM-baryon
displacement. The systems where the halo is starting to fall into the
host ($\mu\approx-1$, $D_{\rm off}>0.6$) should not present a
detectable DM-baryon displacement.

We clearly expect such displacement in the case where the
sub-structure has already passed through the center of the host
halo. In this section we attempt an estimation for the DM-baryon
displacement statistics. We work under the following
hypothesis. First, we consider that systems with $|\mu|<0.9$ have a
baryonic displacement, $d_{2D}^{\rm bar}$, equal to zero. Second, we
consider that all systems with infalling velocities $\mu<-0.9$ and
large displacements $D_{\rm off}>0.6$ also have baryonic displacements
equal to zero. Third, for all the other cases we estimate the
displacement between the baryons and the dominant DM peak by the
distances between the dominant peak and the center of mass of the main
halo, $d_{2D}^{\rm bar}=D_{\rm off} R_{\rm vir}$, where $X_{\rm off}$
is the offset computed for each host halo in the original catalog.

@@

The results for the integrated distributions for $>d_{2D}^{\rm bar}$
are shown in Figure \ref{fig:baryonic_displacements}. The dashed lines
represented the results for groups and the continuous lines correspond
to clusters. As a test of our approach we compare the cluster results
against the analytic fit provided by \cite{ForeroRomero2010}. This fit
reproduces the statistics for the DM-baryon separation found for
clusters more massive than $>10^{14}$\hMsun in a simulation
that included a description for gas with $8$ times the volume of the
Bolshoi Simulation. This fit is valid for separations larger than
$70$\hkpc, beyond that separation we find that it provides a good
description within a factor of $\sim 2$ of

@@

In this Letter we estimate fraction of galaxy groups
that can present observational features associated to a
bullet-like event. This is motivated by the recent observational
results of \citep{Gastaldello} where a system (\bullg) on the mass range
$1\times 10^{14}$\hMsun and velocity dispersion $650$\kms was reported
to feature a displacement between its baryonic (hot gas) and dark
matter components.

Our estimate is based on large N-body dark matter only cosmological
simulation. We estimate the distribution of projected displacements
between the dominant DM clumps in two kinds of systems; groups with
circular velocities $300\kms<V_{\rm c}<700\kms$ and clusters with $V_{\rm
  c}>700$\kms. We report these results at four different redshifts
$z=0.0,0.25,0.5$ and $10$.

We find that a fraction of $50\%$-$60\%$ of the halos in the group
sample present displacement equal or larger than the observed displacement
for \bullg. For halos in the cluster sample this fraction increases to
$80\%$-$85\%$.

We also describe the collision geometry by the angle between
the position and velocity vectors of the sub-halo with respect to the
host. This allows us to separate three different instances: the
sub-halo crossing the host virial radius, the sub-halo going
through the host's center and the secondary infall after reaching
apogee.

From this analysis we derive an estimate for the displacement between
the DM and the baryonic component. In the group sample
$0.1\%$-$1.0\%$ of the halos show a displacement equal or larger than 
the  measurements of \bullg by \citep{Gastaldello}. In the cluster
sample this fraction rises to $4\%$-$10\%$.

The results we present here can be used as a potential test of
$\Lambda$CDM. [...]

>>>>

As a zero-th order approximation, in this paper we work with three
quantities that are available from observations of Bullet-like
systems. The projected distance between two dominant DM clumps, the
projected distance between the DM-baryonic clumps and the ratio of the
mass associated to the DM clumps.

From the simulation point of view, the first quantity can be
translated into the 2D projected values of $||\vec{r}||$ and its value
relative to the virial radius $D_{\rm   off}= ||\vec{r}||_{\rm
  2D}/R_{\rm vir}$. The last quantity, the mass ratio, can be
approximated by ratio of the circular velocities of the two clumps $V_{\rm c,sub}/V_{\rm  c,host}$ which should be equal to the square root of the mass
ratio. The second quantity, the projected DM-baryon distance, is not
directly available from a DM-only simulation but, as we show in the
Results section, can be inferred from the available information.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.9\textwidth]{figure_3.eps}
\end{center}
\caption{Integrated probability distribution for the relative velocity of the
sub-halo with respect to its host. The left panel shows the results
in physical units while the right panel show the same values
normalized by the circular velocity of the host halo. The line encoding
follows the same structure as Figure \ref{fig:displacement}}
\label{fig:velocities}
\end{figure*}

In Figure \ref{fig:velocities} we present the integrated probability
of the relative peculiar velocities of the sub-halos with respect to
the host halo. The left panel presents this velocity in physical units
while the right panel presents them as a fraction of the circular
velocity of the host halo.

The panel with the normalized velocities shows that the distribution
of sub-halo velocities is close to universal. Regardless of the mass
of the host halo and the redshift the integrated distributions lie
very close to each other.  The universality of this profile extends to
the group mass range the findings of \citep{Hayashi2006} for clusters.

@@

Figure \ref{fig:geometry} presents the geometry of the bullet groups using the 
variables $\mu$ and $D_{\rm off}$. The first evident feature is that
most of the configurations have $|mu|>0.9$ ($\theta\leq 30^{\circ}$),
meaning that most of the collisions can be described as a head-on
encounter while only a minority with $|mu|<0.9$ have grazing
trajectories. For the pairs on radial trajectories there are three
regions of interest in this plane that describe different stages in
the collision, assuming that the sub-halo merges (or falls below the
BDM detection threshold) right at its second pass through the center
of the host halo \citep{Poole2006}.

The first region has $\mu\approx-1$ and $D_{\rm off}>0.6$, which
locates the systems where a head-on collision has just started. The
sub-halo is close to the boundary of the host halo and is
infalling. The second region has $\mu\approx 1$ and $D_{\rm off}<0.6$;
at this stage the collision continues after the first crossing of the
host's center, the low number of halos with radial infalling velocities
and displacements $D_{\rm off}>0.6$ suggest that this the maximum
range of radii for the apogee.  The third region corresponds to
$\mu\approx-1$ and $D_{\rm off}<0.6$ which corresponds to the
secondary infall after apogee. We use this sequence in the next
subsection to estimate the expected displacement between baryons and
the dominant DM clump.

@@

Strictly speaking, the results we have derived so far, apply to multi-modal
groups and their expected separation between the two dominant dark
matter clumps. These displacements cannot be interpreted that all have a
corresponding DM-baryon displacement.

However, these different stages for a collision, that we describe at
the end of the previous sub-section, can produce different results in
terms of the displacement between dark matter and baryons. For
instance, the systems where the halo is starting to fall into the host
($\mu\approx-1$, $D_{\rm off}>0.6$) should not present a detectable
DM-baryon displacement.  We do expect such displacements when the
sub-structure has already passed through the center of the host halo,
i.e. cases where $\mu\approx 1$.

In this sub- section we estimate DM-baryon displacement statistics. We
work under the following hypothesis. First, we consider that systems
with $|\mu|<0.9$ have a baryonic displacement, $d_{2D}^{\rm bar}$,
equal to zero. This means that only head-on encounter produce a
displacement. Second, we consider that all systems with infalling
velocities $\mu<-0.9$ and large displacements $D_{\rm off}>0.6$ also
have baryonic displacements equal to zero. Third, for all the other
cases we estimate the displacement between the baryons and the
dominant DM peak by the distances between the dominant peak and the
center of mass of the main halo, $d_{2D}^{\rm bar} = X_{\rm off} R_{\rm
  vir}$, where $X_{\rm off}$ is the offset computed between the minium
of potential and the center of mass for each host halo.

@@

The results for the integrated distributions for $>d_{2D}^{\rm bar}$
are shown in Figure \ref{fig:baryonic_displacements}. The dashed lines
represented the results for groups and the continuous lines correspond
to clusters. As a test of our approach we compare the cluster results
against the analytic fit provided by \cite{ForeroRomero2010}. This fit
reproduces the statistics for the DM-baryon separation found for
clusters more massive than $>10^{14}$\hMsun in a simulation
that included a description for gas with $8$ times the volume of the
Bolshoi Simulation. The fit is valid for separations larger than
$70$\hkpc, beyond which we find that it provides a remarkably good
description within a factor of $\sim 2$ of our results. This gives us
confidence in our approach to estimate the expected fraction of
groups with a DM-baryon displacement.

From Figure \ref{fig:baryonic_displacements} we see that only a
fraction of $0.1\%$ to $1\%$ of the groups are expected to have a
DM-baryon displacement equal or larger than $87\pm 14$\hkpc as
observed in \bullg. This fraction rises to $4\%$ to $10\%$ in the case
of clusters, consistent with the results reported by
\cite{ForeroRomero2010}.

\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{figure_5.eps}
\end{center}
\caption{Integrated probability distribution for the estimated
  baryonic displacements in the group and cluster samples. Continous (dashed)
  lines correspond to clusters (groups). The continuous black line
  marked as $P_{2D}$ shows the statistics reported by
  \citet{ForeroRomero2010} for a cosmological simulation including DM
  and baryons. The vertical lines correspond to the mean value and
  uncertainty of the displacement measured for \bullg.} 
\label{fig:baryonic_displacements}
\end{figure}

@@

In this Letter we estimate fraction of galaxy groups and clusters that
can present observational features associated to a bullet-like
event. This is motivated by the recent observational results of
\citep{Gastaldello} where a system (\bullg) on the mass range $1\times
10^{14}$\hMsun and velocity dispersion $650$\kms was reported to
feature a displacement between its baryonic (hot gas) and dark matter
components.

We estimate the distribution of projected displacements
between the dominant DM clumps in two kinds of systems; groups with
circular velocities $300\kms<V_{\rm c}<700\kms$ and clusters with $V_{\rm
  c}>700$\kms. We report these results at four different redshifts
$z=0.0,0.25,0.5$ and $10$. Our results based on large N-body dark
matter only cosmological simulation with such a resolution that allows
to study the group mass range for the first time in the context of
Bullet-like configurations.

Our main results is that that a fraction of $50\%$-$60\%$ of the halos
in the group sample present displacement equal or larger than the
observed displacement for \bullg. For halos in the cluster sample this
fraction increases to $80\%$-$85\%$. We also derive an
estimate for the displacement between the DM and the baryonic
component. In the group sample $0.1\%$-$1.0\%$ of the halos show a
displacement equal or larger than  the  measurements of \bullg by
\citep{Gastaldello}. In the cluster sample this fraction rises to
$4\%$-$10\%$.

We also find distributions for the DM separation and the velocity of
the bullet through its host. If these quantities are normalized by the
virial radius and the circular velocity, respectively, we arrive at
distributions close to universal that are similar for the two halo
samples at all redshifts.

For the case of \bullg the fair comparison is achieved in the cluster
sample, which have statistics dominated by objects of similar mass. In
this case we conclude that the existence of such configuration is
highly probable in $\Lambda$CDM (4$\%$ to $10\%$ abundance). In turn,
for the same separation the fraction for groups is lower (0.1$\%$ to
$1\%$). Taking into account that the difference in spatial abundance between
these two samples is on the order of a factor of $10$, one can
conclude that the absolute number of groups and clusters presenting a DM-baryon
displacement larger or equal to \bullg should be of the same order.

However, an interesting observational possibility opens up with
surveys such as SL2S that can target a large number of groups and
estimate its multi-modal nature from lensing analysis that can be used
as a potential test of $\Lambda$CDM. In this case the absolute number
of groups with a displacement of the same order of \bullg is larger by
a factor of $\sim 8$ than the number of clusters with the same
displacement.
